---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r set_global_options, echo=FALSE, results='hide'}
Sys.setlocale("LC_TIME", "English")
options(warn=-1)
```

## Loading and preprocessing the data

Load data from .CSV file:

```{r load_and_preprocess}
activity_data <- read.csv("activity.csv", na.strings="NA", 
                 colClasses=c("numeric", "Date", "numeric"))
colnames(activity_data) <- c("Steps", "Date", "Interval")
```

## What is mean total number of steps taken per day?

Code for prepearing data for analysis:

```{r prepare_total_steps_data}
total_steps_data <- aggregate(Steps ~ Date, data=activity_data, sum)
```

Let's plot histogram of the total number of steps taken each day:

```{r plot_total_steps_data}
library(ggplot2)
g <- ggplot(total_steps_data, aes(Date, Steps))
g + geom_bar(stat="identity") +
    ggtitle("Total number of steps taken each day")
```

Next, calculate the mean and median total number of steps taken per day:

```{r calc_mean_median}
mean_steps <- mean(total_steps_data$Steps)
median_steps <- median(total_steps_data$Steps)
paste("Mean =", mean_steps, ", median =", median_steps)
```

## What is the average daily activity pattern?

Code for prepearing data for analysis:

```{r prepare_daily_activity_data}
daily_activity_data <- aggregate(Steps ~ Interval, data=activity_data, mean)
```

Let's make a time series plot:

```{r plot_daily_activity}
g <- ggplot(daily_activity_data, aes(Interval, Steps))
g + geom_line() +
    ylab("Average number of steps") +
    ggtitle("Average number of steps taken for intervals, averaged across all days")
```

Next, find 5-minute interval, contains the maximum number of steps:

```{r find_max_interval}
paste("Interval with maximum steps is", daily_activity_data[which.max(daily_activity_data$Steps), "Interval"])
```

## Imputing missing values

Let's calculate the total number of missing values in the dataset:

```{r calc_missed_val}
paste("Total number of rows with missing values:", sum(!complete.cases(activity_data)))
```

Create a new dataset with filled missing values of steps with mean for that 5-minute interval across all days (using daily_activity_data dataset calculated earlier):

```{r fill_missed_val}
tidy_activity_data <- activity_data
for (i in seq_along(tidy_activity_data[,1])) {
    if (is.na(tidy_activity_data[i, "Steps"])) {
        tidy_activity_data[i, "Steps"] <- as.integer(
            daily_activity_data[daily_activity_data$Interval == tidy_activity_data[i, "Interval"], "Steps"])
    }
}
```

Let's plot histogram of the total number of steps taken each day from new dataset and print new mean and median values of steps:

```{r plot_new_daily_activity}
total_steps_data <- aggregate(Steps ~ Date, data=tidy_activity_data, sum)

library(ggplot2)
g <- ggplot(total_steps_data, aes(Date, Steps))
g + geom_bar(stat="identity") +
    ggtitle("Total number of steps taken each day")

mean_steps <- mean(total_steps_data$Steps)
median_steps <- median(total_steps_data$Steps)
paste("Mean =", mean_steps, ", median =", median_steps)
```

## Are there differences in activity patterns between weekdays and weekends?

Add new factor variable which indicates whether a given date is a weekday or weekend day:

```{r add_weekday_factor}
weekend_col <- weekdays(tidy_activity_data$Date) %in% c("Sunday", "Saturday")
tidy_activity_data$Weekday <- factor(weekend_col, labels=c("weekday", "weekend"))
```

Let's make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days:

```{r plot_panel_weekdays}
plot_data <- aggregate(Steps ~ Interval + Weekday, data=tidy_activity_data, sum)

library(ggplot2)
g <- ggplot(plot_data, aes(Interval, Steps, group=1))
g + geom_line() +
    facet_grid(Weekday ~ .) +
    ggtitle("Activity pattern for weekend and weekdays")
```
